---
title: "Buccal Swab Project qPCR Analysis"
author: "Quinn Le"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())

setwd('/Users/quinnmle/Documents/Castellani Lab/Buccal_2025-12-12/Buccal Swab')
```

# Load Packages
```{r, include=FALSE}
library(tidyverse)
library(ggplot2)
library(magrittr)
library(readxl)
library(data.table)
library(dplyr)
library(raster)
library(cowplot)
```

# Loading in data
```{r}
dlp <- read.csv("dlp_buccal.csv")
alb <- read.csv("alb_buccal.csv")

removed_samples = as.data.frame(matrix(nrow = 1, ncol = 3))
colnames(removed_samples) = c('Sample', 'Well', 'Reason')
dlp$Dlp_CT = dlp$Cq
alb$Alb_CT = alb$Cq
df = merge(dlp, alb, by = c('Well', 'Sample'))
df$Dlp_CT <- as.numeric(as.character(df$Dlp_CT))
df$Alb_CT <- as.numeric(as.character(df$Alb_CT))
```

# Check water samples
```{r}
subset(df, Sample == "Water")$Alb_CT
subset(df, Sample == "Water")$Dlp_CT

df

# plot_grid(
#   subset(df, Sample == "H") %>%
#     ggplot(., aes(y = as.numeric(Alb_CT), x = Sample)) + geom_boxplot() + geom_point() + ylab("CT_ALB"),
#   subset(df, Sample == "Water") %>%
#     ggplot(., aes(y = as.numeric(Dlp_CT), x = Sample)) + geom_boxplot() + geom_point() + ylab("CT_DLP"),
#   nrow = 1, ncol = 2)

```


# Calculating Delta Ct
```{r}
df$deltaCT = df$Alb_CT - df$Dlp_CT

plot_grid(ggplot(df, aes(x = Alb_CT %>% as.numeric())) + 
            geom_density(alpha = 0.5) + xlab('ALB CT') + ylab('Density') + 
            geom_rug() + theme_bw(),
          ggplot(df, aes(x = Dlp_CT %>% as.numeric())) + 
            geom_density(alpha = 0.5) + xlab('DLP CT') + ylab('Density') + 
            geom_rug() + theme_bw(),
          ggplot(df, aes(x = deltaCT %>% as.numeric())) + 
            geom_density(alpha = 0.5) + xlab('Delta CT') + ylab('Density') + 
            geom_rug() + theme_bw(),
          nrow = 1, ncol = 3)

# plot_grid(ggplot(merged_df, aes(x = Alb_CT %>% as.numeric())) + 
#             geom_density(alpha = 0.5) + xlab('ALB CT') + ylab('Density') + 
#             geom_rug() + theme_bw(),
#           ggplot(merged_df, aes(x = Dlp_CT %>% as.numeric())) + 
#             geom_density(alpha = 0.5) + xlab('DLP CT') + ylab('Density') + 
#             geom_rug() + theme_bw(),
#           ggplot(merged_df, aes(x = deltaCT %>% as.numeric())) + 
#             geom_density(alpha = 0.5) + xlab('Delta CT') + ylab('Density') + 
#             geom_rug() + theme_bw(),
#           nrow = 1, ncol = 3)

```

# QC
```{r}
data_for_filtering <- df
print("n to move forward for filtering, including controls:")
nrow(data_for_filtering)
print("n to move forward for filtering, no controls:")
subset(data_for_filtering, Sample != "Water") %>% nrow()
data_for_filtering <- subset(data_for_filtering, Sample != "Water")

```

## Samples with NA DLP CT or ALB CT
```{r}
DLP_ALB_CT_NA_sample <- subset(data_for_filtering, is.na(Dlp_CT) | is.na(Alb_CT))
print("n samples with NA DLP_CT or ALB_CT")
nrow(DLP_ALB_CT_NA_sample)
```

## Samples with DLP CT greater than 26
```{r}
DLP_over26_sample <- subset(data_for_filtering, Dlp_CT > 26)
print("n samples with DLP_CT > 26")
nrow(DLP_over26_sample)
```

## Samples with ALB CT more than 3 SD in any direction
```{r}
highALB <- mean(data_for_filtering$Alb_CT) + (3* (sd(data_for_filtering$Alb_CT)))
lowALB <- mean(data_for_filtering$Alb_CT) - (3* (sd(data_for_filtering$Alb_CT)))

highALB_sample <- subset(data_for_filtering, Alb_CT > highALB)
print("highALB samples")
nrow(highALB_sample)

lowALB_sample <- subset(data_for_filtering, Alb_CT < lowALB)
print("lowALB samples")
nrow(lowALB_sample)

lowALB_sample$qc_assessment <- c('ALB CT < 3 SD from the mean')
#highALB_sample$qc_assessment <- c('ALB CT > 3 SD from the mean')
```

## Samples with deltaCT more than 3 SD in any direction
```{r}
highDCT <- mean(data_for_filtering$deltaCT) + (3*(sd(data_for_filtering$deltaCT)))
lowDCT <- mean(data_for_filtering$deltaCT) - (3*(sd(data_for_filtering$deltaCT)))

highDCT_sample <- subset(data_for_filtering, deltaCT > highDCT)
print("high deltaCT samples")
nrow(highDCT_sample)

lowDCT_sample <- subset(data_for_filtering, deltaCT < lowDCT)
print("low deltaCT samples")
nrow(lowDCT_sample)

lowDCT_sample$qc_assessment <- c('deltaCT < 3 SD from the mean')
highDCT_sample$qc_assessment <- c('deltaCT > 3 SD from the mean')
```

## Removing samples
```{r}
data_for_filtering <- subset(data_for_filtering, Dlp_CT <= 26)
print("n samples after removing >26 DLP CT")
subset(data_for_filtering, Dlp_CT <= 26) %>% nrow()
after_df <- subset(data_for_filtering, Dlp_CT <= 26)

data_for_filtering <- subset(data_for_filtering, Alb_CT > lowALB & Alb_CT < highALB)
print("n samples after removing extreme ALB")
subset(data_for_filtering, Alb_CT > lowALB & Alb_CT < highALB) %>% nrow()
after_df <- subset(after_df, Alb_CT > lowALB & Alb_CT < highALB)

data_for_filtering <- subset(data_for_filtering, deltaCT > lowDCT & deltaCT < highDCT)
print("n samples after removing extreme delta CT")
subset(data_for_filtering, deltaCT > lowDCT & deltaCT < highDCT) %>% nrow()
after_df <- subset(after_df, deltaCT > lowDCT & deltaCT < highDCT)
```

## Replot after QC
```{r}
plot_grid(ggplot(after_df, aes(x = Alb_CT %>% as.numeric())) + 
            geom_density(alpha = 0.5) + xlab('ALB CT') + ylab('Density') + 
            geom_rug() + theme_bw(),
          ggplot(after_df, aes(x = Dlp_CT %>% as.numeric())) + 
            geom_density(alpha = 0.5) + xlab('DLP CT') + ylab('Density') + 
            geom_rug() + theme_bw(),
          ggplot(after_df, aes(x = deltaCT %>% as.numeric())) + 
            geom_density(alpha = 0.5) + xlab('Delta CT') + ylab('Density') + 
            geom_rug() + theme_bw(),
          nrow = 1, ncol = 3)
```

## Summary of low quality samples to be removed
```{r}
print("Samples with ALB CT > 3 SD in any direction")
lowALB_sample$Sample 
highALB_sample$Sample

print("Samples with DLP CT > 26")
DLP_over26_sample$Sample

print("Samples with delta CT > 3 SD in any direction")
lowDCT_sample$Sample
highDCT_sample$Sample
```
# Loading in age and sex data
```{r}
anno_df <- read.csv("Quinn_Limo_SampleIDs.csv")
colnames(anno_df) <- c("Sample", "Limo.ID", "Sex", "Age")
anno_df$Sex <- as.factor(anno_df$Sex)
anno_df$Sample <- as.character(anno_df$Sample)
merged_df <- left_join(after_df, anno_df, by = "Sample")


png("buccal_AgeSexPlots_20251212.png")
p_age <- ggplot(merged_df, aes(x = Age)) +
  geom_density(na.rm = TRUE) +
  geom_rug(sides = "b", alpha = 0.4, na.rm = TRUE) +
  geom_vline(aes(xintercept = mean(Age, na.rm = TRUE)),
             linetype = "dashed") +
  labs(x = "Age", y = "Density") +
  theme_classic()

p_sex <- ggplot(merged_df, aes(x = factor(Sex))) +
  geom_bar(na.rm = TRUE) +
  labs(x = "Sex", y = "Count") +
  theme_classic()

plot_grid(p_age, p_sex, nrow = 1)

anno_df$Age
```

# Adjusting for age and sex
```{r}
merged_df$Sex <- as.factor(merged_df$Sex)
fit <- lm(deltaCT ~ Age + Sex, data = merged_df)
summary(fit)
res <- resid(fit)
merged_df$deltaCT_adj <- scale(res)
sd(merged_df$deltaCT_adj)
mean(merged_df$deltaCT_adj)
head(merged_df$deltaCT_adj)

# mitochondrialdnacop adjusted
```

# Generating reports
```{r}
# _minimal: only sample name and CT values
# _CLEAN: removal of samples that failed QC

write.csv(df, "Buccal_qPCR_Results_20251212.csv", quote=F)
write.csv(merged_df, "Buccal_qPCR_Results_20251212_CLEAN.csv", quote=F)

## Edit
merged_df$Castellani.ID <- merged_df$Sample
merged_df$mtDNACN_adj_age_sex <- merged_df$deltaCT_adj
merged_df$mtDNACN_unadj <- scale(merged_df$deltaCT)
sd(merged_df$deltaCT_standardized)

## Edit/

df2 <- df[,c("Sample", "deltaCT", "Alb_CT","Dlp_CT")]
df2_clean <- merged_df[,c("Castellani.ID", "Limo.ID", "deltaCT", "Alb_CT","Dlp_CT", "mtDNACN_adj_age_sex", "Age", "Sex")]
df2_final <- merged_df[,c("Castellani.ID", "Limo.ID", "deltaCT", "Alb_CT","Dlp_CT", "mtDNACN_unadj", "mtDNACN_adj_age_sex", "Age", "Sex")]

write.csv(df2, "Buccal_qPCR_Results_20251212_minimal.csv", quote=F)
write.csv(df2_clean, "Buccal_qPCR_Results_20251212_CLEAN_minimal.csv", quote=F)

write.csv(df2_final, "Buccal_qPCR_Results_20251212_outliersremoved_minimal_final.csv", quote=F)

forPlot <- df2[order(df2$Sample),]
forPlot_clean <- df2_clean[order(df2_clean$Sample),]

data <- forPlot
for (i in 1:nrow(data)){
  data$type[i]=data$Sample[i]
}
data_clean <- forPlot_clean
for (i in 1:nrow(data_clean)){
  data_clean$type[i]=data_clean$Sample[i]
}

data$SampleID <- as.numeric(data$Sample)
data <- arrange(data,SampleID)
data$type <- factor(data$type, levels=data$SampleID)

data$Dlp_CT <- as.numeric(data$Dlp_CT)
data$Alb_CT <- as.numeric(data$Alb_CT)
minC <- min(data$deltaCT)
minD <- min(data$Dlp_CT)
minA <- min(data$Alb_CT)

data_clean$SampleID <- as.numeric(data_clean$Sample)
data_clean <- arrange(data_clean,SampleID)
data_clean$type <- factor(data_clean$type, levels=data_clean$SampleID)

data_clean$Dlp_CT <- as.numeric(data_clean$Dlp_CT)
data_clean$Alb_CT <- as.numeric(data_clean$Alb_CT)
minC <- min(data_clean$deltaCT)
minD <- min(data_clean$Dlp_CT)
minA <- min(data_clean$Alb_CT)

png(file = "buccal_deltaCt_20251212.png", width = 1500, height = 400)
ggplot(data) + geom_boxplot(aes(col=type, x=type, y=deltaCT)) +  ggtitle("buccal_deltaCt_20251212") + ylab("deltaCt") + theme(legend.title=element_blank(), axis.title = element_text(face="bold")) + geom_point(aes(col=type, x=type, y=deltaCT), position=position_dodge(0.75)) +  theme(axis.text.x=element_text(angle = 45, hjust = 1), plot.margin = margin(1, 0.2, 1.5, 2, "cm"), text = element_text(size=10), axis.title.x=element_blank(), legend.title=element_blank()) + ylim(minC-0.2, minC+5.5)
dev.off()

png(file = "buccal_dlp_20251212.png", width = 1500, height = 400)
ggplot(data) + geom_boxplot(aes(col=type, x=type, y=Dlp_CT)) +  ggtitle("buccal_dlp_20251212") + ylab("Dlp_CT") + theme(legend.title=element_blank(), axis.title = element_text(face="bold")) + geom_point(aes(col=type, x=type, y=Dlp_CT), position=position_dodge(0.75)) +  theme(axis.text.x=element_text(angle = 45, hjust = 1), plot.margin = margin(1, 0.2, 1.5, 2, "cm"), text = element_text(size=10), axis.title.x=element_blank(), legend.title=element_blank()) + ylim(minD-0.2,minD+5.5)
dev.off()

png(file = "buccal_alb_20251212.png", width = 1500, height = 400)
ggplot(data) + geom_boxplot(aes(col=type, x=type, y=Alb_CT)) +  ggtitle("buccal_alb_20251212") + ylab("Alb_CT") + theme(legend.title=element_blank(), axis.title = element_text(face="bold")) + geom_point(aes(col=type, x=type, y=Alb_CT), position=position_dodge(0.75)) +  theme(axis.text.x=element_text(angle = 45, hjust = 1), plot.margin = margin(1, 0.2, 1.5, 2, "cm"), text = element_text(size=10), axis.title.x=element_blank(), legend.title=element_blank()) + ylim(minA-0.2, minA+5.5)
dev.off()

png(file = "buccal_deltaCt_adj_20251212_CLEAN.png", width = 1500, height = 400)
ggplot(data_clean) + geom_boxplot(aes(col=type, x=type, y=deltaCT_adj)) +  ggtitle("buccal_deltaCt_adj_20251212_CLEAN") + ylab("deltaCt_adj") + theme(legend.title=element_blank(), axis.title = element_text(face="bold")) + geom_point(aes(col=type, x=type, y=deltaCT_adj), position=position_dodge(0.75)) +  theme(axis.text.x=element_text(angle = 45, hjust = 1), plot.margin = margin(1, 0.2, 1.5, 2, "cm"), text = element_text(size=10), axis.title.x=element_blank(), legend.title=element_blank()) + ylim(-4,4)
dev.off()

png(file = "buccal_deltaCt_20251212_CLEAN.png", width = 1500, height = 400)
ggplot(data_clean) + geom_boxplot(aes(col=type, x=type, y=deltaCT)) +  ggtitle("buccal_deltaCt_20251212_CLEAN") + ylab("deltaCt") + theme(legend.title=element_blank(), axis.title = element_text(face="bold")) + geom_point(aes(col=type, x=type, y=deltaCT), position=position_dodge(0.75)) +  theme(axis.text.x=element_text(angle = 45, hjust = 1), plot.margin = margin(1, 0.2, 1.5, 2, "cm"), text = element_text(size=10), axis.title.x=element_blank(), legend.title=element_blank()) + ylim(minC-0.2, minC+5.5)
dev.off()

png(file = "buccal_dlp_20251212_CLEAN.png", width = 1500, height = 400)
ggplot(data_clean) + geom_boxplot(aes(col=type, x=type, y=Dlp_CT)) +  ggtitle("buccal_dlp_20251212") + ylab("Dlp_CT") + theme(legend.title=element_blank(), axis.title = element_text(face="bold")) + geom_point(aes(col=type, x=type, y=Dlp_CT), position=position_dodge(0.75)) +  theme(axis.text.x=element_text(angle = 45, hjust = 1), plot.margin = margin(1, 0.2, 1.5, 2, "cm"), text = element_text(size=10), axis.title.x=element_blank(), legend.title=element_blank()) + ylim(minD-0.2,minD+5.5)
dev.off()

png(file = "buccal_alb_20251212_CLEAN.png", width = 1500, height = 400)
ggplot(data_clean) + geom_boxplot(aes(col=type, x=type, y=Alb_CT)) +  ggtitle("buccal_alb_20251212") + ylab("Alb_CT") + theme(legend.title=element_blank(), axis.title = element_text(face="bold")) + geom_point(aes(col=type, x=type, y=Alb_CT), position=position_dodge(0.75)) +  theme(axis.text.x=element_text(angle = 45, hjust = 1), plot.margin = margin(1, 0.2, 1.5, 2, "cm"), text = element_text(size=10), axis.title.x=element_blank(), legend.title=element_blank()) + ylim(minA-0.2, minA+5.5)
dev.off()
```

